---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---


## Loading and preprocessing the data
The data is loaded from the csv file, and two fields are added:
one for the day of the week and a flag to indicate whether it
is a weekend day.
``` {r load, results='hide'}
library(dplyr)
actall <- read.csv("activity.csv")
actall$date <- as.Date(actall$date, format="%Y-%m-%d")
#Add fields for day of the week and the weekend flag
mutate(actall, actall$day <- weekdays(date, abbreviate=TRUE))
mutate(actall, actall$weekend <- ifelse(day %in% c("Sat", "Sun"), "y", "n"))
act <- filter(actall, is.na(steps) == FALSE)

bydate <- group_by(act, date)
bydate <- summarize(bydate, sum(steps))
names(bydate) <- c("date", "totalsteps")
```



## What is mean total number of steps taken per day?
 

``` {r Mean}
hist(bydate$totalsteps, xlab = "Total Steps", main = "Histogram of Total Steps\n(missing data ignored)", col = "red")
bydateo <- bydate
print(summary(bydate$totalsteps))
#m <- filter(bydate, bydate$averagesteps == max(bydate$averagesteps))
```

### The mean total number of steps taken per day is: `r as.integer(mean(bydate$totalsteps))` 


## What is the average daily activity pattern?
``` {r daily}
bydate <- group_by(act, interval)
bydate <- summarise(bydate, round(mean(steps), digits=1))
names(bydate) <- c("interval", "averagesteps")

plot(bydate, type = "l", main = "Average Daily Activity Pattern\n(missing data ignored)", ylab = "Average Steps", col = "blue")
m <- filter(bydate, bydate$averagesteps == max(bydate$averagesteps))
print(summary(bydate))
```
### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?: `r m$interval`


## Imputing missing values
### There are `r nrow(subset(actall, is.na(actall$steps) == TRUE))` rows in the table with missing steps value.

Now we replace all NA steps values for each interval with the average with that interval for the 
data set.
``` {r Impute}
acto <- act
act <- actall
# Replace NA steps with average for that interval across the table
for(i in unique(act$interval))
{
        x <- filter(act, interval == i & is.na(steps) == FALSE)
        act <- mutate(act, steps = ifelse(is.na(steps) == TRUE & interval == i, round(mean(x$steps), digits=1), steps))
}
```

# Compare the graphs between omitting NAs and imuting the missing values

Histogram differences:

``` {r }
bydate <- group_by(act, date)
bydate <- summarize(bydate, sum(steps))
names(bydate) <- c("date", "totalsteps")
par(mfrow = c(2, 1))
hist(bydate$totalsteps, xlab = "Total Steps", main = "Histogram of Total Steps\n(missing data immuted)", col = "red")
bydateo <- group_by(acto, date)
bydateo <- summarize(bydateo, sum(steps))
names(bydateo) <- c("date", "totalsteps")
hist(bydateo$totalsteps, xlab = "Total Steps", main = "Histogram of Total Steps\n(missing data ignored)", col = "red")
```


### The mean total number of steps taken per day is (top graph): `r as.integer(mean(bydate$totalsteps))` 
### The mean total number of steps taken per day is (bottom graph): `r as.integer(mean(bydateo$totalsteps))` 
Activity Pattern Differences:
```{r}
par(mfrow = c(2, 1))
bydate <- group_by(act, interval)
bydate <- summarise(bydate, round(mean(steps), digits=1))
names(bydate) <- c("interval", "averagesteps")

plot(bydate, type = "l", main = "Average Daily Activity Pattern\n(missing data immuted)", ylab = "Average Steps", col = "blue")
m <- filter(bydate, bydate$averagesteps == max(bydate$averagesteps))
bydateo <- group_by(acto, interval)
bydateo <- summarise(bydateo, round(mean(steps), digits=1))
names(bydateo) <- c("interval", "averagesteps")

plot(bydateo, type = "l", main = "Average Daily Activity Pattern\n(missing data ignored)", ylab = "Average Steps", col = "blue")
mo <- filter(bydateo, bydateo$averagesteps == max(bydateo$averagesteps))
```


The interval with max steps for top graph: `r as.integer(m$interval)` vs bottom graph `r as.integer(mo$interval)`
indicates that with a low frequency of NA steps, ignoring the NA values, or imuting them with the average
makes no difference in the calculated values when expressing them as integers.


## Are there differences in activity patterns between weekdays and weekends?
We can see from the following plots that on weekdays there is one major peak in the morning,
but during the weekend there are several peaks throughout the day
```{r compare}
par(mfrow = c(2, 1))
# Ave for weekend only
actw <- filter(act, weekend=="y")
bydate <- group_by(actw, interval)
bydate <- summarise(bydate, round(mean(steps), digits=1))
plot(bydate, type = "l", main = "Average Daily Activity Pattern\nWeekend Days", ylab = "Average Steps", col = "blue")
print(summary(bydate))
# Ave for non weekend only
actw <- filter(act, weekend=="n")
bydate <- group_by(actw, interval)
bydate <- summarise(bydate, round(mean(steps), digits=1))
plot(bydate, type = "l", main = "Non Weekend Days", ylab = "Average Steps", col = "blue")

print(summary(bydate))
```


